|

### 2023-05-13 paper/03-methods.Rmd
Edit methods notation and update results description

Moreover, since we observe STRAIN continuously at every tenth of a second within each play, we can then compute the average STRAIN across all frames played for every pass rusher.
Formally, the average STRAIN, denoted by $\overline{\text{STRAIN}}$, for pass rusher $i$ over $n$ total plays with $T$ total frames is
[CHECK THIS - is notation consistent with what we have above?]
\begin{equation}
\overline{\text{STRAIN}}_{i} = \frac{1}{n} \sum_{t=1}^T {\text{STRAIN}}_{it}.
\end{equation}
This can be helpful for player evaluation, as we determine the most effective pass-rushers based on their average STRAIN values in Section \ref{sec:strainbar}.
We also use average STRAIN to assess statistical properties of our metric in Section \ref{sec:statprop}.

## Multilevel Model

@@

While average STRAIN is a simple first step for evaluating the pass rush, the observed average STRAIN on a single play is likely due to a convolution of factors.
Outside of the pass rusher's ability, there is variability in the opposing strength of pass blockers across plays a pass rusher is involved.
Additionally, based on previous work by PFF, the opposing QB plays an important role on observed pressure in passing plays [@eager2019pressure].
Thus, we need to appropriately divide the credit of an observed average STRAIN across the different players involved and other factors.

To this end, we fit a multilevel model to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their pass blocker, QB, and play-level attributes.
We identify the pass blocker linked with the pass rusher of interest using the data provided by PFF.
Since there can be multiple blockers linked to a rusher, for simplicity, we consider the nearest blocker positioned to the pass rusher at the start of the play.
We use random intercepts for the three player groups: (1) pass rushers as $R$, (2) nearest pass blockers as $B$, and (3) QBs as $Q$.
The coefficients $\beta$ for additional information $X_{ij}$ about pass rusher $j$ in play $i$ are estimated as fixed-effects. 
Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows.
\begin{equation}
\begin{aligned}
\overline{\text{STRAIN}}_{ij} &\sim N(R_{r[ij]} + B_{b[ij]} + Q_{q[i]} + X_{ij} \cdot \beta, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
R_{r} &\sim N(\mu_R, \sigma^2_R), \text{ for } r = 1, \cdots, \text{ \# of pass rushers}, \\
B_{b} &\sim N(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
Q_{q} &\sim N(\mu_Q, \sigma^2_Q), \text{ for } q = 1, \cdots, \text{ \# of QBs}.
\end{aligned}
\end{equation}

We consider a normal distribution to obtain the random intercepts for each player toward their respective position group means.
This provides us with average individual effects on STRAIN for each player while applying shrinkage, which is a useful property in this problem's context since we do not observe the same number of plays for each player.
With regards to the fixed-effects about pass rusher $j$ in play $i$, we include the length of the play (i.e. the elapsed time between the ball snap and a QB event), the positions of pass rusher and nearest blocker, and the number of pass blockers assigned to the pass rusher based on the PFF data.
Table \ref{tab:pos} shows our positional categorization for the pass rush and pass block roles.
We implement the model using penalized likelihood via the `lme4` package in `R` [@bateslme4; @R2023Language].

>>>>

Moreover, since we observe STRAIN continuously at every tenth of a second within each play, we can then compute the average STRAIN across all frames played for every pass rusher.
Formally, the average STRAIN, denoted by $\overline{\text{STRAIN}}$, for pass rusher $j$ involved in $n_j$ total plays across $T_j$ total frames is

\begin{equation}
\overline{\text{STRAIN}}_{j} = \frac{1}{T_j} \sum_{t=1}^{T_j} \widehat{\text{STRAIN}}_{j}(t).
\end{equation}
This can be helpful for player evaluation, as we determine the most effective pass-rushers based on their average STRAIN values in Section \ref{sec:strainbar}.
We also use average STRAIN to assess statistical properties of our metric in Section \ref{sec:statprop}.

## Multilevel model of play-level STRAIN

@@

While average STRAIN is a simple first step for evaluating the pass rush, we can also measure pass rusher $j$'s observed average STRAIN on a single play $i$ consisting of $T_i$ total frames,

\begin{equation}
\overline{\text{STRAIN}}_{ij} = \frac{1}{T_i} \sum_{t=1}^{T_i} \widehat{\text{STRAIN}}_{ij}(t).
\end{equation}

However, the observed average STRAIN on a single play is likely due to a convolution of factors.
Outside of the pass rusher's ability, there is variability in the opposing strength of pass blockers across plays a pass rusher is involved.
Additionally, based on previous work by PFF, the opposing QB plays an important role on observed pressure in passing plays [@eager2019pressure].
Thus, we need to appropriately divide the credit of an observed average STRAIN across the different players involved and other factors.

To this end, we fit a multilevel model to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their pass blocker, QB, and play-level attributes.
We identify the pass blocker linked with the pass rusher of interest using the data provided by PFF.
Since there can be multiple blockers linked to a rusher, for simplicity, we consider the nearest blocker positioned to the pass rusher at the start of the play.
We use random intercepts for the three player groups: (1) pass rushers as $R$, (2) nearest pass blockers as $B$, and (3) QBs as $Q$.
The coefficients $\boldsymbol{\beta}$ for additional information vector of information  $\mathbf{x_{ij}}$ about pass rusher $j$ in play $i$ are estimated as fixed-effects. 
Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows.
\begin{equation}
\begin{aligned}
\overline{\text{STRAIN}}_{ij} &\sim N(R_{j[i]} + B_{b[ij]} + Q_{q[i]} + \mathbf{x_{ij}} \cdot \boldsymbol{\beta}, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
R_{j} &\sim N(\mu_R, \sigma^2_R), \text{ for } j = 1, \cdots, \text{ \# of pass rushers}, \\
B_{b} &\sim N(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
Q_{q} &\sim N(\mu_Q, \sigma^2_Q), \text{ for } q = 1, \cdots, \text{ \# of QBs}.
\end{aligned}
\end{equation}

We consider a normal distribution to obtain the random intercepts for each player toward their respective position group means.
This provides us with average individual effects on STRAIN for each player while applying shrinkage, which is a useful property in this problem's context since we do not observe the same number of plays for each player.
With regards to the fixed-effects about pass rusher $j$ in play $i$, we include the length of the play (i.e., the elapsed time between the ball snap and a QB event), the positions of pass rusher and nearest blocker, and the number of pass blockers assigned to the pass rusher based on the PFF data.
Table \ref{tab:pos} shows our positional categorization for the pass rush and pass block roles.
We implement the model using penalized likelihood via the `lme4` package in `R` [@bateslme4; @R2023Language]. Additionally, we generate estimates of uncertainty for the player-level random intercepts based on 95% confidence intervals generated by sampling from the posterior distribution via parametric bootstrapping with the `merTools` package [@knowlesmerTools, @gelman2006data].

### 2023-05-13 paper/04-results.Rmd
Edit methods notation and update results description

In addition to adjusting for opposition, our multilevel model allows us to provide some notion of uncertainty about the pass rusher's effect on the play level STRAIN.
Figure \ref{fig:fig_rankings_boot} displays the mean varying intercept for the top ten pass rushers for each position along with 95% confidence intervals generated by sampling from the posterior distributions via parametric bootstrapping using the `merTools` package [@knowlesmerTools, @gelman2006data]. 
We observe similar rankings based on the random intercepts in comparison to the average STRAIN results in Tables \ref{tab:edge} and \ref{tab:interior}, with Myles Garrett (DE), Aaron Donald (DT), Vita Vea (NT), and T.J. Watt (OLB) as the top pass rusher for their respective position. 
We observe different levels of variability across these top players, with the only a subset of players with intervals strictly above zero.

[NEED TO THINK ABOUT WHAT MORE TO SAY HERE]

>>>>

In addition to adjusting for opposition, our multilevel model allows us to provide some notion of uncertainty about the pass rusher's effect on the play level STRAIN.
Figure \ref{fig:fig_rankings_boot} displays the mean varying intercept for the top ten pass rushers for each position along with 95% confidence intervals. 
We observe similar rankings based on the random intercepts in comparison to the average STRAIN results in Tables \ref{tab:edge} and \ref{tab:interior}, with Myles Garrett (DE), Aaron Donald (DT), Vita Vea (NT), and T.J. Watt (OLB) as the top pass rusher for their respective position. 
We observe different levels of variability across these top players, with the only a subset of players with intervals strictly above zero. This is not necessarily surprising given the limited sample of data. Additionally, we are ignoring correlated performance between pass rushers on the same team, e.g., INSERT TWO PLAYERS ON SAME TEAM, but leave exploration of this for future work.
